---
title: "Trees as Point Patterns"
author: "Danny Foster"
date: "12/27/2021"
output: html_document
---

Using the `spatstat` package to analyze the spatial distribution of trees 
from crystal cave road as realizations of point pattern processes.

# Setup

Load packages and data from SEKI:

```{r}
library(here)
library(tidyverse)
library(spatstat)

treelist = 
  readRDS(here::here('02-data', '04-geolocated', 'seki_trees.rds')) %>%
  # left join in the metadata to get the plot centers
  left_join(
    readRDS(here::here('02-data', '04-geolocated', 'seki_metadata.rds')) %>%
      select(plot_id, x_center, y_center, mort) %>%
      group_by(plot_id, mort, x_center, y_center) %>%
      summarise() %>%
      ungroup()
  ) %>%
  # and get the within-plot coordinates of each tree
  mutate(x_rel = x_coord-x_center,
         y_rel = y_coord-y_center)



summary(treelist)

trees = 
  treelist %>%
  filter(dbh_cm >= 11.4)

saplings = 
  treelist %>%
  filter(dbh_cm < 11.4)

```

# Data munging

`spatstat` has its own set of data objects that it uses, so step one is 
to convert the trees data (which is a treelist with coordinates) into 
a set of point patterns (a "hyperframe") and construct a "window" object to 
represent the plot area. 

```{r}
# rejects 5 trees which occur right on the plot boundary, which is OK 
trees.solist = 
  lapply(X = unique(trees$plot_id)[order(unique(trees$plot_id))],
         FUN = function(p){
           trees_in_plot = trees %>% filter(plot_id==p)
           trees_as_ppp = 
             ppp(x = trees_in_plot$x_rel,
                 y = trees_in_plot$y_rel,
                 marks = trees_in_plot %>% select(dbh_cm, status),
                 # plot area, constructed of two 30x10m rectangles
                 window = as.owin(layered(as.owin(c(-15, 15, -5, 5)),
                                        as.owin(c(-5, 5, -15, 15)))))
         }) %>%
  # convert to a list of spatial objects
  as.solist()

saplings.solist = 
  lapply(X = unique(saplings$plot_id)[order(unique(saplings$plot_id))],
         FUN = function(p){
           saplings_in_plot = saplings %>% filter(plot_id==p)
           saplings_as_ppp = 
             ppp(x = saplings_in_plot$x_rel,
                 y = saplings_in_plot$y_rel,
                 marks = saplings_in_plot %>% select(dbh_cm, status),
                 # plot area, constructed of two 30x10m rectangles
                 window = as.owin(layered(as.owin(c(-15, 15, -1, 1)),
                                        as.owin(c(-1, 1, -15, 15)))))
         }) %>%
  # convert to a list of spatial objects
  as.solist()

trees_plot_info = 
  trees %>%
  group_by(plot_id, mort) %>%
  summarise() %>%
  ungroup()

saplings_plot_info = 
  saplings %>%
  group_by(plot_id, mort) %>%
  summarise() %>%
  ungroup()

#plot the point pattern for a single plot
plot(trees.solist[[1]])
plot(saplings.solist[[1]])


trees.hyperframe = 
  hyperframe(pattern = trees.solist,
             plot_id = trees_plot_info$plot_id,
             mort = trees_plot_info$mort)

saplings.hyperframe = 
  hyperframe(pattern = saplings.solist,
             plot_id = saplings_plot_info$plot_id,
             mort = saplings_plot_info$mort)


# plot the point pattern for all plots (can't get the marks to show here)
plot(trees.hyperframe$pattern, use.marks = FALSE)

plot(saplings.hyperframe$pattern, use.marks = FALSE)

```

Little weird that its plotting them at slightly different zooms.

# Basic data exploration

Just using some of the builtin summary functions from spatstat:

```{r}
summary(trees.hyperframe$pattern)
summary(saplings.hyperframe$pattern)
```

We can also plot the predicted underlying intensity kernel for each windows. 
These images assume the range parameter for the underlying intensity function 
is 3.75m (1/8th the window range). 

```{r}
plot(density(trees.hyperframe$pattern))
```

The package includes several methods of estimating the true bandwidth, which 
overlaps with our GP work on surface fuels.

Here are intensity surfaces using the DBHs as weights:

```{r}
plot(density(trees.hyperframe$pattern$`1`,
             weights = trees.hyperframe$pattern$`1`$marks$dbh_cm))
```

# Basic measures of interpoint correlation

We can estimate ripley's $K$ ($K(r)$ is the number of neighbors expected 
within distance $r$ for a pattern) for each plot, and compare it to ripley's 
$K$ for spatially random patterns (no interpoint interaction):


```{r}
lapply(X = 1:length(trees.solist),
       FUN= function(i){
         plot(envelope(trees.solist[[i]], Kest, nsim = 99, fix.n = TRUE))
       })
lapply(X = 1:length(saplings.solist),
       FUN= function(i){
         plot(envelope(saplings.solist[[i]], Kest, nsim = 99, fix.n = TRUE))
       })

```

Most of the plots' $K$ functions (black lines) are within the envelope of 
plausible $K$ values generated by completely random simulations (grey bounds), 
though there are a few that exhibit "significant" clustering or randomness. 
(This sort of scanning for deviations from the null hypothesis reeks of 
data dredging; a more formal analysis would call for pre-specifying a 
specific range of distances we want to test.) 

This exploration suggests that the within-plot distribution of trees is not
inconsistent with randomness. 
This would be convenient (simulation is easy!) but boring, and doesn't seem 
very plausible ecologically (there are ecological processes like 
reproduction and competition which we would expect to generate clustering 
or dispersion, respectively). We're also interested in knowing whether the 
process generating the spatial distribution of trees varies across classes of 
drought mortality. 

More of the sapling patterns exhibit clustering.

# Using the USGS demography plots

## Comparing plot locations with drought mortality levels

```{r}
library(here)
library(sf)
library(raster)
library(tidyverse)

usgs_plots = 
  st_read(here::here('02-data',
                       '00-source',
                       'usgs',
                       'Boundary Polygons.gdb'),
            layer = 'allplotscombined') %>%
  
  st_intersection(
    st_read(here::here('02-data',
                       '00-source',
                       'seki_dx',
                       'dxstudyarea.shp'))
  )

usgs_plots

mort_categories = 
  st_read(here::here('02-data',
                     '02-intermediate',
                     'intermediate_plot_location_files',
                     'mort_categories.shp'))


library(tmap)
tmap_mode('view')
tm_shape(mort_categories)+
  tm_polygons('mort')+
  tm_shape(usgs_plots)+
  tm_borders('black')


library(RODBC)

db = 
  odbcConnectAccess2007(here::here('02-data', 
                               '00-source',
                               'usgs',
                               'DemogDATABASE2021_10_08.accdb'))



usgs = 
  sqlFetch(db, 'allplotarchive') %>%
  
  # keep only the two plots from the DX AOI
  filter(is.element(PLOT, unique(usgs_plots$PLOT)))

usgs %>%
  group_by(PLOT, YEAR6, YEAR7) %>%
  summarise()
# CCRPIPO was visited in 2015/2016 and 2020/2021
# BBBPIPO was visited in 2017

```


# A Gibbs model for tree occurrences

## Background and theory

Notes from the `spatstat` textbook by Baddley et al., chapter 13. I'll bounce 
back to the earlier material in chapters 9 and 10 as neeeded. 

### 13.2 conditional intensity

A Poisson point process is completely specified by its intensity function 
$\lambda(u)$. The number of occurrences within some area is a Poisson-distributed 
draw from the underlying mean, which can vary across space as some 
linear combination of fixed and/or random effects. The resulting point pattern 
may exhibit clustering if the intensity surface is not homogeneous. This allows 
them to represent e.g. tree density across areas of varying suitability. Other 
processes leading to clustering, such as reproduction, can be represented 
with Cox and/or cluster models. However, 
the Poisson model assumes that the occurrences themselves are independent, and 
cannot capture processes resulting from direct interaction between trees, such 
as competition. A Gibbs point process, on the other hand, can represent direct 
interaction between trees. It does so by making the intensity surface for 
each tree conditional on the observed locations of all the other trees. 
Basically, the presence of a tree changes the underlying intensity surface 
(e.g. you can represent competition by making the intensity lower in a 
neighborhood around existing trees). 