---
title: "Proposed Models"
author: "Danny Foster"
date: "3/14/2022"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(cmdstanr)
library(posterior)
library(bayesplot)
```


# Litter and/or duff

## Model Description

The model for litter and/or duff depths is:

$$Y_i \sim \text{NegBinomial}(\mu_i, \kappa)$$

where:

  - $Y_i$ is the observed depth of litter and/or duff at observation $i$ in 
  centimeters, rounded to the nearest cm
  - $\mu_i$ is the unobserved true mean depth at observation $i$
  - $\kappa$ is the negative binomial dispersion parameter, here assumed constant 
  across all samples
  
$$\text{log}(\mu_i) = \textbf{X}\beta + \textbf{GP}_{l,p} + b_p$$

where:

  - $\textbf{X}$ is the design matrix of fixed effects covariates
  - $\beta$ are the coefficients for the fixed effects
  - $\textbf{GP}$ is a matrix of realized gaussian process random effects, $L$ with rows 
  for the (fixed across plots) spatial locations $l = 1,2,...,L$ and $P$ columns 
  for the plots p$p = 1,2,...P$. $l$ is the index of the location for the 
  observation $i$ and $p$ is the index for the plot of of the observation $i$.
  - $b$ is a vector of normally distributed random effects for the plots, 
  $b \sim \text{Normal}(0,\sigma_{plot}^2)$, and $b_p$ is the random effect 
  for plot $p$, the plot associated with observation $i$.
  
$$\textbf{GP}_{1:L,p} \sim \text{MVN}(0,\Sigma)$$

where:

  - $\Sigma$ is the $L$ by $L$ covariance matrix of sample locations within 
  a plot. The gaussian process parameters, and thus the covariance matrix, 
  are constant within a group of plots. For clarity, we present models for 
  only a single group. For analyses explicitly comparing multiple fuel 
  models, various parameters (including $\beta$,
  $\Sigma$, $\alpha$ and $\rho$) will be allowed to vary by group. 
  
$$\Sigma_g = \alpha_g^2 \text{exp}(\frac{-D^2}{2\rho_g^2})$$

where:

  - $\alpha_g$ is the standard deviation of the gaussian process random effect for 
  group $g$, controlling the strength of the spatial effect
  - $D^2$ is the $L$ by $L$ matrix of squared distances among all sample locations 
  $l = 1,...,L$ in a plot
  - $\rho_g$ is the length-scale parameter of the gaussian process random effect 
  for group $g$, 
  controlling the rate at which the correlation between observations decays 
  as the distance between them increases


## Simulate Sample Data

```{r}
set.seed(112188)

#### data structure ############################################################
G = 1 # 1 group
L = 64 # 64 subsamples within each plot
P = 10 # 10 plots
N = L*P # total number of observations

# within-plot sample locations, coordinates in meters relative to plot center
sample_coords = 
  data.frame(x_coord = 
               c(-29.5, -19.5, -9.5, -7.5, -7, -5.5, -4.5, -4, -3.5, -3, 
                 -2.5, -2.4, -2.3, -2.2, -2.1, -2, 2, 2.1, 2.2, 2.3, 2.4, 2.5,
                 3, 3.5, 4, 4.5, 5.5, 7, 7.5, 9.5, 19.5, 29.5,
                 rep(0, times = 32)),
             y_coord = 
               c(rep(0, times = 32), 
                 -29.5, -19.5, -9.5, -7.5, -7, -5.5, -4.5, -4, -3.5, -3, 
                 -2.5, -2.4, -2.3, -2.2, -2.1, -2, 2, 2.1, 2.2, 2.3, 2.4, 2.5,
                 3, 3.5, 4, 4.5, 5.5, 7, 7.5, 9.5, 19.5, 29.5))

# vector of location IDs for within-plot sample locations
location_id = 
  rep(x = 1:L, times = P)

# vector of plot IDs for within-plot sample locations
plot_id = 
  sapply(X = 1:P,
         FUN = function(p){
           rep(p, times = L)
         }) %>%
  as.numeric()

# vector of group IDs for each plot
group_id = rep(1, times = P)

# design matrix for fixed effects
X = 
  data.frame(
    intercept = rep(1, times = N)
  ) %>%
  as.matrix()



#### parameters ################################################################


# fixed effects coefficients; here group means are the only fixed effects
#beta = rnorm(n = G, mean = 0, sd = 2) 
beta = c(2.5)
  
  
# standard deviation of the plot random effect
#sigmaPlot = rnorm(n = 1, mean = 0, sd = 2)
sigmaPlot = 0.25


# negative binomial dispersion parameter
#kappa = rcauchy(n = 1, location = 0, scale = 5)
#while (kappa<=0){
#  kappa = rcauchy(n = 1, location = 0, scale = 5)
#}
kappa = 1



# standard deviation of the GP random effect for group g
#alpha = truncnorm::rtruncnorm(n = G, a = 0, b = Inf, mean = 0, sd = 0.5)
alpha = c(0.5)


# length scale of the GP random effect for group g
#rho = invgamma::rinvgamma(n = G, shape = 5, rate = 40)
rho = c(4)


#### random effects realizations ###############################################

# matrix of within-plot pairwise distances
D = dist(sample_coords) %>% as.matrix()

# within-plot covariance matrices, one for each group
K = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = 
          sapply(X = 1:G,
                 FUN = function(g){
                   
                   sapply(X = 1:L,
                          FUN = function(l){
                            
                            sapply(X = 1:L,
                                   FUN = function(lprime){
                                     #paste0('G:',g,'L:',l,'Lp:',lprime)
                                     alpha[g]**2 * 
                                       exp(-(D[lprime,l]**2)/
                                             (2*rho[g]**2))
                                   })
                          })
                   
                 }))

# add a small noise term to ensure positive definiteness
for (g in 1:G){
  K[,,g] = K[,,g]+diag(x = 1e-9, nrow = L, ncol = L)
}

# cholesky decompositions of the covariance matrices, one for each group
KL = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = sapply(X = 1:G,
                      FUN = function(g){
                        
                        t(chol(K[,,g]))
                        
                      }))

# standard normal deviates for spatial random effects
zGP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = rnorm(n = L*P, mean = 0, sd = 1))

# realized spatial random effects
GP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = 
           sapply(X = 1:P,
                  FUN = function(p){
                    KL[,,group_id[p]] %*% zGP[,p]
                  }))

# realized plot random effects
plot_effects = sigmaPlot * rnorm(n = P, mean = 0, sd = 1)

#### linear predictors and responses ###########################################

logmu = 
  # fixed effects
  X %*% beta +
  # spatial random effect
  sapply(X = 1:N,
         FUN = function(i){GP[location_id[i],plot_id[i]]}) +
  
  # plot random effect
  sapply(X = 1:N,
         FUN = function(i){plot_effects[plot_id[i]]})

Y = rnbinom(n = N, size = kappa, mu = exp(logmu))

hist(Y)
ggplot(data = data.frame(Y),
       aes(x = Y))+
  geom_histogram()+
  scale_x_continuous(limits = c(0, 100))

```


## Estimate Parameters

Assemble model data object and compile stan model:

```{r}

data.litterduff = 
  list(N = N,
       P = P,
       L = L,
       G = G,
       J = G,
       
       Y = Y,
       plot_id = plot_id,
       location_id = location_id,
       group_id = group_id,
       
       X = X,
       coords = sample_coords)


stan_model.litterduff = cmdstan_model(here::here('03-analysis',
                                      'litterduff_depths.stan'))

```

Estimate parameters:

```{r}

litterduff_fit = stan_model.litterduff$sample(data = data.litterduff,
                                   parallel_chains = 4,
                                   output_dir = 
                                     here::here('02-data', '06-results', 'demo'),
                                   output_basename = 'litterduff',
                                   seed = 112188,
                                   adapt_delta = 0.9)


```

Model diagnostics:

```{r}

litterduff_fit$cmdstan_diagnose()

litterduff_fit$summary(c('beta', 'alpha', 'rho',
                         'sigmaPlot', 'kappa')) %>%
  print(n = Inf)


litterduff_fit$save_object(here::here('02-data', '06-results', 'demo', 
                                      'litterduff_fit.rds'))

```

Posterior distributions vs. real values:

```{r}
litterduff_draws = litterduff_fit$draws(
                         variables = c('alpha', 'rho','beta',
                                                     'sigmaPlot', 'kappa')) %>%
  as_draws_df()

head(litterduff_draws)

ggplot(data = litterduff_draws,
       aes(x = `alpha[1]`))+
  geom_histogram()+
  geom_vline(xintercept = alpha,
             color = 'red')+
  theme_minimal()

ggplot(data = litterduff_draws,
       aes(x = `rho[1]`))+
  geom_histogram()+
  geom_vline(xintercept = rho,
             color = 'red')+
  theme_minimal()

ggplot(data = litterduff_draws,
       aes(x = `beta[1]`))+
  geom_histogram()+
  geom_vline(xintercept = beta,
             color = 'red')+
  theme_minimal()


ggplot(data = litterduff_draws,
       aes(x = kappa))+
  geom_histogram()+
  geom_vline(xintercept = kappa,
             color = 'red')+
  theme_minimal()

ggplot(data = litterduff_draws,
       aes(x = sigmaPlot))+
  geom_histogram()+
  geom_vline(xintercept = sigmaPlot,
             color = 'red')+
  theme_minimal()

```

Looks good.


# Woody Debris


## Model Description

There are two submodels in the woody debris model. The first models the 
tally of woody debris particles intersecting the transect, and is very similar 
to the litter/duff model. The second submodel models the diameter of the 
woody debris particles using a gamma distribution. 

### Particle Tallies

The model for woody debris tallies is:

$$Y_i \sim \text{NegBinomial}(\mu_i, \kappa)$$

where:

  - $Y_i$ is the observed count of woody debris particles (1h, 10h, 100h, 1000h) 
  at observation $i$, which is a 1m subtransect 
  - $\mu_i$ is the unobserved true mean count at observation $i$
  - $\kappa$ is the negative binomial dispersion parameter, here assumed constant 
  across all samples

$$\text{log}(\mu_i) = \textbf{X}\beta + \textbf{GP}_{l,p} + b_p + z_t$$

where:

  - $\textbf{X}$ is the design matrix of fixed effects covariates
  - $\beta$ are the coefficients for the fixed effects
  - $\textbf{GP}$ is a matrix of realized gaussian process random effects, $L$ with rows 
  for the (fixed across plots) spatial locations $l = 1,2,...,L$ and $P$ columns 
  for the plots p$p = 1,2,...P$. $l$ is the index of the location for the 
  observation $i$ and $p$ is the index for the plot of of the observation $i$.
  - $b$ is a vector of normally distributed random effects for the plots, 
  $b \sim \text{Normal}(0,\sigma_{plot}^2)$, and $b_p$ is the random effect 
  for plot $p$, the plot associated with observation $i$.
  - $z$ is a vector of normally distributed random effects for the transects,
  $z \sim \text{Normal}(0,\sigma_{transect}^2)$, and $z_t$ is the random effect 
  for transect $t$, the transect associated with observation $i$.
  
$$\textbf{GP}_{1:L,p} \sim \text{MVN}(0,\Sigma)$$

where:

  - $\Sigma$ is the $L$ by $L$ covariance matrix of sample locations within 
  a plot. The gaussian process parameters, and thus the covariance matrix, 
  are constant within a group of plots. For clarity, we present models for 
  only a single group. For analyses explicitly comparing multiple fuel 
  models, various parameters (including $\beta$,
  $\Sigma$, $\alpha$ and $\rho$) will be allowed to vary by group. 
  
$$\Sigma_g = \alpha_g^2 \text{exp}(\frac{-D^2}{2\rho_g^2})$$

where:

  - $\alpha_g$ is the standard deviation of the gaussian process random effect for 
  group $g$, controlling the strength of the spatial effect
  - $D^2$ is the $L$ by $L$ matrix of squared distances among all sample locations 
  $l = 1,...,L$ in a plot
  - $\rho_g$ is the length-scale parameter of the gaussian process random effect 
  for group $g$, 
  controlling the rate at which the correlation between observations decays 
  as the distance between them increases



### Particle Diameters

The model for woody particle diameters is:

$$Y_i \sim \text{InvGamma}(\alpha_d, \beta_d)$$

where:

  - $Y_i$ is the log of the particle diameter in mm
  - $\alpha_d$ is the scale parameter for the diameter model
  - $\beta_d$ is the rate parameter for the diameter model

## Simulate Sample Data

Woody tallies:

```{r}
set.seed(112188)

#### data structure ############################################################
G = 1 # 1 group
L = 32 # 64 subsamples within each plot
P = 10 # 10 plots
NT = P*2 # total number of transects
N = L*P # total number of observations

# within-plot sample locations, coordinates in meters relative to plot center
sample_coords = 
  data.frame(x_coord = 
               c(-29.5, -19.5, -9.5, -7.5, -5.5, -4.5,  -3.5,  -2.5, 2.5,
                  3.5, 4.5, 5.5, 7.5, 9.5, 19.5, 29.5,
                 rep(0, times = 16)),
             y_coord =  
               c(rep(0, times = 16),
                 -29.5, -19.5, -9.5, -7.5, -5.5, -4.5,  -3.5,  -2.5, 2.5,
                  3.5, 4.5, 5.5, 7.5, 9.5, 19.5, 29.5))

# vector of location IDs for within-plot sample locations
location_id = 
  rep(x = 1:L, times = P)

# vector of plot IDs for within-plot sample locations
plot_id = 
  sapply(X = 1:P,
         FUN = function(p){
           rep(p, times = L)
         }) %>%
  as.numeric()

# vector of transect IDs for within plot sample locations
transect_id = 
  sapply(X = 1:NT,
         FUN = function(t){rep(t, times = 16)}) %>%
  as.numeric()

# vector of group IDs for each plot
group_id = rep(1, times = P)

# design matrix for fixed effects
X = 
  data.frame(
    intercept = rep(1, times = N)
  ) %>%
  as.matrix()



#### parameters ################################################################


# fixed effects coefficients; here group means are the only fixed effects
#beta = rnorm(n = G, mean = 0, sd = 2) 
beta = c(3.5)
  
  
# standard deviation of the plot random effect
#sigmaPlot = rnorm(n = 1, mean = 0, sd = 2)
sigmaPlot = 0.25

sigmaTransect = 0.1

# negative binomial dispersion parameter
#kappa = rcauchy(n = 1, location = 0, scale = 5)
#while (kappa<=0){
#  kappa = rcauchy(n = 1, location = 0, scale = 5)
#}
kappa = 1



# standard deviation of the GP random effect for group g
#alpha = truncnorm::rtruncnorm(n = G, a = 0, b = Inf, mean = 0, sd = 0.5)
alpha = c(0.5)


# length scale of the GP random effect for group g
#rho = invgamma::rinvgamma(n = G, shape = 5, rate = 40)
rho = c(10)


#### random effects realizations ###############################################

# matrix of within-plot pairwise distances
D = dist(sample_coords) %>% as.matrix()

# within-plot covariance matrices, one for each group
K = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = 
          sapply(X = 1:G,
                 FUN = function(g){
                   
                   sapply(X = 1:L,
                          FUN = function(l){
                            
                            sapply(X = 1:L,
                                   FUN = function(lprime){
                                     #paste0('G:',g,'L:',l,'Lp:',lprime)
                                     alpha[g]**2 * 
                                       exp(-(D[lprime,l]**2)/
                                             (2*rho[g]**2))
                                   })
                          })
                   
                 }))

# add a small noise term to ensure positive definiteness
for (g in 1:G){
  K[,,g] = K[,,g]+diag(x = 1e-9, nrow = L, ncol = L)
}

# cholesky decompositions of the covariance matrices, one for each group
KL = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = sapply(X = 1:G,
                      FUN = function(g){
                        
                        t(chol(K[,,g]))
                        
                      }))

# standard normal deviates for spatial random effects
zGP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = rnorm(n = L*P, mean = 0, sd = 1))

# realized spatial random effects
GP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = 
           sapply(X = 1:P,
                  FUN = function(p){
                    KL[,,group_id[p]] %*% zGP[,p]
                  }))

# realized plot random effects
plot_effects = sigmaPlot * rnorm(n = P, mean = 0, sd = 1)

# realized transect random effects
transect_effects = sigmaTransect *rnorm(n = NT, mean = 0, sd = 1)

#### linear predictors and responses ###########################################

logmu = 
  # fixed effects
  X %*% beta +
  # spatial random effect
  sapply(X = 1:N,
         FUN = function(i){GP[location_id[i],plot_id[i]]}) +
  
  # plot random effect
  sapply(X = 1:N,
         FUN = function(i){plot_effects[plot_id[i]]}) +
  
  # transect random effect
  sapply(X = 1:N,
         FUN = function(i){transect_effects[transect_id[i]]})

Y = rnbinom(n = N, size = kappa, mu = exp(logmu))

hist(Y)
ggplot(data = data.frame(Y),
       aes(x = Y))+
  geom_histogram()

data.woody_tallies = 
  list(N = N,
       P = P,
       NT = NT,
       L = L,
       G = G,
       J = G,
       Y = Y,
       plot_id = plot_id,
       location_id = location_id,
       transect_id = transect_id,
       group_id = group_id,
       
       X = X,
       coords = sample_coords)


```

Woody diameters:

```{r}
set.seed(110819)

#### data structure ############################################################

G = 1
P = 10
N = 1800 # similar to the actual data we have
J = G

# vector of plot IDs for within-plot sample locations
plot_id = 
  sapply(X = 1:P,
         FUN = function(p){
           rep(p, times = N/P)
         }) %>%
  as.numeric()

# vector of group IDs for each plot
group_id = rep(1, times = P)

# design matrix for fixed effects
X = 
  data.frame(
    intercept = rep(1, times = N)
  ) %>%
  as.matrix()

#### parameter values ##########################################################

alpha = c(1)
beta = c(2)

#### random effects realizations ###############################################

#### linear predictors and responses ###########################################


Y = 
  sapply(X = 1:N,
         FUN = function(i){
           invgamma::rinvgamma(n = 1, 
                               shape = alpha[group_id[plot_id]], 
                               rate = beta[group_id[plot_id]])
         }) %>%
  as.numeric()

hist(Y)
data.woody_diams = 
  list(N = N,
       P = P,
       G = G,
       J = J,
       X = X,
       
       group_id = group_id,
       plot_id = plot_id,
       
       Y = Y)


```

## Estimate Parameters

Compile stan models:

```{r}



stan_model.woody_tallies = cmdstan_model(here::here('03-analysis',
                                      'woody_tallies.stan'))

stan_model.woody_diams = cmdstan_model(here::here('03-analysis',
                                                  'woody_diams.stan'))

```

Estimate parameters:

```{r}

woody_tallies_fit = stan_model.woody_tallies$sample(data = data.woody_tallies,
                                   parallel_chains = 4,
                                   output_dir = 
                                     here::here('02-data', '06-results', 'demo'),
                                   output_basename = 'woodytallies',
                                   seed = 112188)

woody_diams_fit = 
  stan_model.woody_diams$sample(data = data.woody_diams,
                                parallel_chains = 4,
                                output_dir = 
                                  here::here('02-data', '06-results', 'demo'),
                                output_basename = 'woodydiams',
                                seed = 110819)

```

Model diagnostics:

```{r}

woody_tallies_fit$cmdstan_diagnose()

woody_tallies_fit$summary(c('beta', 'alpha', 'rho',
                         'sigmaPlot', 'kappa')) %>%
  print(n = Inf)


woody_tallies_fit$save_object(here::here('02-data', '06-results', 'demo', 
                                      'woody_tallies_fit.rds'))

```

Posterior distributions vs. real values:

```{r}
woody_tallies_draws = woody_tallies_fit$draws(
                         variables = c('alpha', 'rho','beta',
                                                     'sigmaPlot', 
                                       'sigmaTransect','kappa')) %>%
  as_draws_df()

head(woody_tallies_draws)

ggplot(data = woody_tallies_draws,
       aes(x = `alpha[1]`))+
  geom_histogram()+
  geom_vline(xintercept = alpha,
             color = 'red')+
  theme_minimal()

ggplot(data = woody_tallies_draws,
       aes(x = `rho[1]`))+
  geom_histogram()+
  geom_vline(xintercept = rho,
             color = 'red')+
  theme_minimal()

ggplot(data = woody_tallies_draws,
       aes(x = `beta[1]`))+
  geom_histogram()+
  geom_vline(xintercept = beta,
             color = 'red')+
  theme_minimal()


ggplot(data = woody_tallies_draws,
       aes(x = kappa))+
  geom_histogram()+
  geom_vline(xintercept = kappa,
             color = 'red')+
  theme_minimal()

ggplot(data = woody_tallies_draws,
       aes(x = sigmaPlot))+
  geom_histogram()+
  geom_vline(xintercept = sigmaPlot,
             color = 'red')+
  theme_minimal()

```

Looks good.


```{r}

woody_diams_fit$cmdstan_diagnose()

woody_diams_fit$summary(c('alpha', 'beta')) %>%
  print(n = Inf)


woody_diams_fit$save_object(here::here('02-data', '06-results', 'demo', 
                                      'woody_diams_fit.rds'))

```

Posterior distributions vs. real values:

```{r}
woody_diams_draws = woody_diams_fit$draws(
                         variables = c('alpha', 'beta')) %>%
  as_draws_df()

head(woody_diams_draws)


ggplot(data = woody_diams_draws,
       aes(x = `alpha[1]`))+
  geom_histogram()+
  geom_vline(xintercept = alpha,
             color = 'red')+
  theme_minimal()

ggplot(data = woody_diams_draws,
       aes(x = `beta[1]`))+
  geom_histogram()+
  geom_vline(xintercept = beta,
             color = 'red')+
  theme_minimal()


```

Looks good.

# Understory Vegetation

## Model Description

The understory vegetation model treats understory vegetation as binomial 
distribution, where the number of trials is the number of horizontal 0.25m 
layers in which vegetation could be present, and the number of successes is 
the number of layers (counting up from the ground) in which vegetation was 
actually present. 

$$Y_i \sim \text{Binomial}(M,\eta_i)$$

where:

  - $Y_i$ is the number of layers in which vegetation is present at sample $i$,
  $0 \leq Y_i \leq M$
  - $M$ is the maximum number of layers in which understory vegetation was 
  tested (maximum height of 3.5m divided into 0.25m segments = 14 layers, so 
  $M=14$ for all locations
  - $\eta_i$ is the per-layer probability of presence
  
$\eta$ is linked to a linear predictor via a logit link:  

$$\text{logit}(\eta) = \textbf{X}\beta_H + \textbf{GP}_H + b_H$$

With the design matrix $\textbf{X}$, fixed effects coefficients $\beta$,
gaussian process random effects $\textbf{GP}$, and plot-level random effects 
$b$ all as defined above.

## Simulate Sample Data

```{r}
set.seed(110819)

#### data structure ############################################################
G = 1 # 1 group
L = 60 # 64 subsamples within each plot
P = 10 # 10 plots
N = L*P # total number of observations
M = 14 # maximum number of layers (counting up from the ground) in which vegetation is present


# within-plot sample locations, coordinates in meters relative to plot center
sample_coords = 
  data.frame(x_coord = 
               c(seq(-14.5, 14.5, 1),
                 rep(0, times = 30)),
             y_coord =  
               c(rep(0, times = 30),
                 seq(-14.5, 14.5, 1)))

# vector of location IDs for within-plot sample locations
location_id = 
  rep(x = 1:L, times = P)

# vector of plot IDs for within-plot sample locations
plot_id = 
  sapply(X = 1:P,
         FUN = function(p){
           rep(p, times = L)
         }) %>%
  as.numeric()

# vector of group IDs for each plot
group_id = rep(1, times = P)

# design matrix for fixed effects
X = 
  data.frame(
    intercept = rep(1, times = N)
  ) %>%
  as.matrix()



#### parameters ################################################################


# fixed effects coefficients; here group means are the only fixed effects
#beta = rnorm(n = G, mean = 0, sd = 2) 
beta = c(0.5)
  
# standard deviation of the plot random effect
#sigmaPlot = rnorm(n = 1, mean = 0, sd = 2)
sigmaPlot = 0.1

# standard deviation of the GP random effect for group g
#alpha = truncnorm::rtruncnorm(n = G, a = 0, b = Inf, mean = 0, sd = 0.5)
alpha = c(1)


# length scale of the GP random effect for group g
#rho = invgamma::rinvgamma(n = G, shape = 5, rate = 40)
rho = c(10)

#### random effects realizations ###############################################

# matrix of within-plot pairwise distances
D = dist(sample_coords) %>% as.matrix()

# within-plot covariance matrices, one for each group
K = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = 
          sapply(X = 1:G,
                 FUN = function(g){
                   
                   sapply(X = 1:L,
                          FUN = function(l){
                            
                            sapply(X = 1:L,
                                   FUN = function(lprime){
                                     #paste0('G:',g,'L:',l,'Lp:',lprime)
                                     alpha[g]**2 * 
                                       exp(-(D[lprime,l]**2)/
                                             (2*rho[g]**2))
                                   })
                          })
                   
                 }))

# add a small noise term to ensure positive definiteness
for (g in 1:G){
  K[,,g] = K[,,g]+diag(x = 1e-9, nrow = L, ncol = L)
}

# cholesky decompositions of the covariance matrices, one for each group
KL = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = sapply(X = 1:G,
                      FUN = function(g){
                        
                        t(chol(K[,,g]))
                        
                      }))

# standard normal deviates for spatial random effects
zGP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = rnorm(n = L*P, mean = 0, sd = 1))

# realized spatial random effects
GP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = 
           sapply(X = 1:P,
                  FUN = function(p){
                    KL[,,group_id[p]] %*% zGP[,p]
                  }))

# realized plot random effects
plot_effects = sigmaPlot * rnorm(n = P, mean = 0, sd = 1)


#### linear predictors and responses ###########################################

# presence absence
logit_eta = 
  # fixed effects 
  X %*% beta +
  # spatial random effect
  sapply(X = 1:N,
         FUN = function(i){GP[location_id[i],plot_id[i]]}) +
  
  # plot random effect
  sapply(X = 1:N,
         FUN = function(i){plot_effects[plot_id[i]]})


Y = 
  rbinom(n = N, size = 14, prob = boot::inv.logit(logit_eta))
  
  

hist(Y)
ggplot(data = data.frame(Y),
       aes(x = Y))+
  geom_histogram()

data.veg = 
  list(N = N,
       M = M,
       P = P,
       L = L,
       G = G,
       J = G,
       Y = Y,
       plot_id = plot_id,
       location_id = location_id,
       group_id = group_id,
       
       X = X,
       coords = sample_coords)


```



## Estimate Parameters

Compile stan models:

```{r}

stan_model.veg = cmdstan_model(here::here('03-analysis',
                                      'veg_heights.stan'))

```

Estimate parameters:

```{r}

veg_fit = stan_model.veg$sample(data = data.veg,
                                   parallel_chains = 4,
                                   output_dir = 
                                     here::here('02-data', '06-results', 'demo'),
                                   output_basename = 'veg',
                                   seed = 110819,
                                adapt_delta = 0.95)

```

Model diagnostics:

```{r}

veg_fit$cmdstan_diagnose()

veg_fit$summary(c('beta', 'alpha', 'rho',
                         'sigmaPlot')) %>%
  print(n = Inf)


veg_fit$save_object(here::here('02-data', '06-results', 'demo', 
                                      'veg_fit.rds'))

```

Posterior distributions vs. real values:

```{r}
veg_draws = veg_fit$draws(
                         variables = c('beta', 'alpha', 'rho',
                         'sigmaPlot')) %>%
  as_draws_df()

head(veg_draws)

ggplot(data = veg_draws,
       aes(x = `alpha[1]`))+
  geom_histogram()+
  geom_vline(xintercept = alpha,
             color = 'red')+
  theme_minimal()

ggplot(data = veg_draws,
       aes(x = `rho[1]`))+
  geom_histogram()+
  geom_vline(xintercept = rho,
             color = 'red')+
  theme_minimal()

ggplot(data = veg_draws,
       aes(x = `beta[1]`))+
  geom_histogram()+
  geom_vline(xintercept = beta,
             color = 'red')+
  theme_minimal()

ggplot(data = veg_draws,
       aes(x = sigmaPlot))+
  geom_histogram()+
  geom_vline(xintercept = sigmaPlot,
             color = 'red')+
  theme_minimal()


```

Looks pretty good.


# Saplings and Trees

## Model Description

## Simulate Sample Data

## Estimate Parameters


# Simulate fuelbeds based on real data

## Litter and duff

Extract posterior median parameter values estimated from actual data:

```{r}
litterduff_real_params = 
  readRDS(here::here('02-data',
                     '06-results',
                     'real_fits',
                     'litterduff_fit.rds'))$draws(
                       
                       variables = c('alpha[1]', 'rho[1]','beta[1]',
                                                     'alpha[2]', 'rho[2]','beta[2]',
                                                     'alpha[3]', 'rho[3]','beta[3]',
                                                     'sigmaPlot', 'kappa')) %>%
  as_draws_df() %>%
  summarise_all(median)
                       
```

Simulate based on these realistic parameter values:

```{r}
set.seed(112188)

#### data structure ############################################################
G = 3 # 2 groups
L = 1156 # 64 subsamples within each plot
P = 3 #
N = L*P # total number of observations

# within-plot sample locations, coordinates in meters relative to plot center
grid_coords_small = 
  expand.grid(x_coord = seq(from = 0, to = 10, by = 0.3),
              y_coord = seq(from = 0, to = 10, by = 0.3))

# vector of location IDs for within-plot sample locations
location_id = 
  rep(x = 1:L, times = P)

# vector of plot IDs for within-plot sample locations
plot_id = 
  sapply(X = 1:P,
         FUN = function(p){
           rep(p, times = L)
         }) %>%
  as.numeric()

# vector of group IDs for each plot
group_id = rep(1:3, times = P/3)

# design matrix for fixed effects
X = 
  data.frame(
    intercept = rep(1, times = N),
    group2 = ifelse(group_id[plot_id]==2,1,0),
    group3 = ifelse(group_id[plot_id]==3,1,0)
  ) %>%
  as.matrix()



#### parameters ################################################################


# fixed effects coefficients; here group means are the only fixed effects
#beta = rnorm(n = G, mean = 0, sd = 2) 
beta = c(litterduff_real_params$`beta[1]`,
         litterduff_real_params$`beta[2]`,
         litterduff_real_params$`beta[3]`)
  
  
# standard deviation of the plot random effect
#sigmaPlot = rnorm(n = 1, mean = 0, sd = 2)
sigmaPlot = litterduff_real_params$sigmaPlot


# negative binomial dispersion parameter
#kappa = rcauchy(n = 1, location = 0, scale = 5)
#while (kappa<=0){
#  kappa = rcauchy(n = 1, location = 0, scale = 5)
#}
kappa = litterduff_real_params$kappa



# standard deviation of the GP random effect for group g
#alpha = truncnorm::rtruncnorm(n = G, a = 0, b = Inf, mean = 0, sd = 0.5)
alpha = c(litterduff_real_params$`alpha[1]`,
          litterduff_real_params$`alpha[2]`,
          litterduff_real_params$`alpha[3]`)


# length scale of the GP random effect for group g
#rho = invgamma::rinvgamma(n = G, shape = 5, rate = 40)
rho = c(litterduff_real_params$`rho[1]`,
        litterduff_real_params$`rho[2]`,
        litterduff_real_params$`rho[3]`)


#### random effects realizations ###############################################

# matrix of within-plot pairwise distances
D = dist(grid_coords_small) %>% as.matrix()

# within-plot covariance matrices, one for each group
K = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = 
          sapply(X = 1:G,
                 FUN = function(g){
                   
                   sapply(X = 1:L,
                          FUN = function(l){
                            
                            sapply(X = 1:L,
                                   FUN = function(lprime){
                                     #paste0('G:',g,'L:',l,'Lp:',lprime)
                                     alpha[g]**2 * 
                                       exp(-(D[lprime,l]**2)/
                                             (2*rho[g]**2))
                                   })
                          })
                   
                 }))

# add a small noise term to ensure positive definiteness
for (g in 1:G){
  K[,,g] = K[,,g]+diag(x = 1e-9, nrow = L, ncol = L)
}

# cholesky decompositions of the covariance matrices, one for each group
KL = 
  array(dim = c(L, L, G),
        dimnames = list('l' = 1:L,
                        'lprime' = 1:L,
                        'g' = 1:G),
        data = sapply(X = 1:G,
                      FUN = function(g){
                        
                        t(chol(K[,,g]))
                        
                      }))

# standard normal deviates for spatial random effects
zGP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = rnorm(n = L*P, mean = 0, sd = 1))

# realized spatial random effects
GP = 
  matrix(nrow = L, ncol = P, byrow = FALSE,
         data = 
           sapply(X = 1:P,
                  FUN = function(p){
                    KL[,,group_id[p]] %*% zGP[,p]
                  }))

# realized plot random effects
plot_effects = sigmaPlot * rnorm(n = P, mean = 0, sd = 1)

#### linear predictors and responses ###########################################

logmu = 
  # fixed effects
  X %*% beta +
  # spatial random effect
  sapply(X = 1:N,
         FUN = function(i){GP[location_id[i],plot_id[i]]}) +
  
  # plot random effect
  sapply(X = 1:N,
         FUN = function(i){plot_effects[plot_id[i]]})

Y = rnbinom(n = N, size = kappa, mu = exp(logmu))

hist(Y)
ggplot(data = data.frame(Y),
       aes(x = Y))+
  geom_histogram()+
  scale_x_continuous(limits = c(0, 100))


sim_grid = 
  X %>%
  as.data.frame() %>%
  mutate(location_id = location_id,
          plot_id = plot_id,
         Y = Y) %>%
  left_join(grid_coords_small %>%
              mutate(location_id = 1:L))

ggplot(data = sim_grid,
       aes(x = x_coord, y = y_coord, fill = Y))+
  geom_tile()+
  scale_fill_viridis_c()+
  theme_minimal()+
  facet_wrap(~plot_id)+
  coord_fixed()+
  labs(fill = 'Litter+duff depth (cm)')

```


